---
title: "ps1"
format: pdf
tables: TRUE 
---

"Initial Project Setup"

GitHub Repository: <https://github.com/meganmartinez-25/Gov51-ps1.git>

#Part 2

1.  

```{r, include = FALSE}
#Loading tidyverse and reading in my ACS data
library(tidyverse)
acs2024_raw <- read.csv("gov51-ps1/data/raw/acs2024.csv")

```

2.  Variables: STATEFIP = State (FIPS code), SEX= (Sex), AGE = (Age), EDUC = Education attainment (general version), EMPSTAT = Employement status (general version), INCTOT = Total Personal Income, TRANTIME = Travel Time to Work. For TRANTIME, 000 signifies N/A. THE PERWT is attempting to describe how many people in the population are represented by each "sample case."

3.  

```{r}

acs2024 <-acs2024_raw %>% 
  select("STATEFIP", "SEX", "AGE", "EDUC", "EMPSTAT","INCTOT","TRANTIME") %>% 
  mutate(female = if_else(SEX == 2, 1, 0), 
  less_than_hs = if_else(EDUC %in% c(1,2), 1, 0),
  hs_only = if_else(EDUC %in% c(3,4,5,6), 1, 0), 
  some_college = if_else(EDUC %in% c(7,8,9), 1, 0), 
  college_only = if_else(EDUC == 10,1,0),
  advanced_degree = if_else(EDUC == 11, 1, 0)) |> 
  
  mutate(employed = if_else(EMPSTAT == 1, 1, 0, 
              if_else(EMPSTAT %in% c(0,9), NA_integer_, 0)),
  unemployed = if_else(EMPSTAT == 2, 1, 0, 
                if_else(EMPSTAT %in% c(0,9), NA_integer_, 0)), 
  not_in_labor_force = if_else(EMPSTAT == 3, 1, 0, 
   if_else(EMPSTAT %in% c(0,9), NA_integer_, 0))) |>
  mutate(INCTOT = if_else(INCTOT %in% c(9999999, 9999998), NA_real_, INCTOT))

#question about the NA_character
#do i need to mutate TRANTIME on this one yet? 

```

4.  If you treated N/A codes as real zeroes when calculating summary statistics, this would skew the data to the left, because for a variable that is not supposed to have any 0 values, if there is even one, with a range from 0 to 888, having zeroes part of the final calculation will cause the mean, or average, to lower which will depict a smaller average commute time in minutes.

```{r}

library(knitr)

acs2024_rename <- acs2024 %>% 
  rename( "Age" = "AGE", 
     "Female" = "female" , 
     "Less than High School"= "less_than_hs" , 
     "High School Only"= "hs_only",
     "Some College" = "some_college" , 
     "College Only"= "college_only" , 
    "Advanced Degree" = "advanced_degree" , 
   "Employed" =  "employed" , 
     "Unemployed" = "unemployed" , 
     "Not in Labor Force" = "not_in_labor_force", 
     "Commute Time (mins)" = "TRANTIME", 
    "Total Income ($)" = "INCTOT")

acs2024_stats <- data.frame(
  Variable = c("Age", "Female", "Less than High School", "High School Only",  "Some College", "College Only", "Advanced Degree", "Employed", "Unemployed", "Not in Labor Force", "Commute Time (mins)", "Total Income ($)"),
  N = c(length(acs2024_rename$`Age`),
        length(acs2024_rename$`Female`),
        length(acs2024_rename$`Less than High School`),
        length(acs2024_rename$`High School Only`),
        length(acs2024_rename$`Some College`),
        length(acs2024_rename$`College Only`),
        length(acs2024_rename$`Advanced Degree`),
        length(acs2024_rename$`Employed`),
        length(acs2024_rename$`Unemployed`),
        length(acs2024_rename$`Not in Labor Force`),
        length(acs2024_rename$`Commute Time (mins)`),
        length(acs2024_rename$`Total Income ($)`)),
  Mean = c(mean(acs2024_rename$`Age`),
           mean(acs2024_rename$`Female`), 
           mean(acs2024_rename$`Less than High School`),
           mean(acs2024_rename$`High School Only`), 
           mean(acs2024_rename$`Some College`),
           mean(acs2024_rename$`College Only`),
           mean(acs2024_rename$`Advanced Degree`), 
           mean(acs2024_rename$`Employed`),
           mean(acs2024_rename$`Unemployed`), 
           mean(acs2024_rename$`Not in Labor Force`),
           mean(acs2024_rename$`Commute Time (mins)`),
           mean(acs2024_rename$`Total Income ($)`, na.rm = TRUE)),
  SD = c(sd(acs2024_rename$`Age`),
         sd(acs2024_rename$`Female`), 
         sd(acs2024_rename$`Less than High School`),
         sd(acs2024_rename$`High School Only`), 
         sd(acs2024_rename$`Some College`),
         sd(acs2024_rename$`College Only`), 
         sd(acs2024_rename$`Advanced Degree`),
         sd(acs2024_rename$`Employed`), 
         sd(acs2024_rename$`Unemployed`),
         sd(acs2024_rename$`Not in Labor Force`), 
         sd(acs2024_rename$`Commute Time (mins)`),
         sd(acs2024_rename$`Total Income ($)`, na.rm=TRUE)),
  Min = c(min(acs2024_rename$`Age`),
          min(acs2024_rename$`Female`), 
          min(acs2024_rename$`Less than High School`),
          min(acs2024_rename$`High School Only`), 
          min(acs2024_rename$`Some College`),
          min(acs2024_rename$`College Only`), 
          min(acs2024_rename$`Advanced Degree`),
          min(acs2024_rename$`Employed`), 
          min(acs2024_rename$`Unemployed`),
          min(acs2024_rename$`Not in Labor Force`), 
          min(acs2024_rename$`Commute Time (mins)`),
          min(acs2024_rename$`Total Income ($)`, na.rm = TRUE)),
  Max = c(max(acs2024_rename$`Age`),
          max(acs2024_rename$`Female`), 
          max(acs2024_rename$`Less than High School`),
          max(acs2024_rename$`High School Only`), 
          max(acs2024_rename$`Some College`),
          max(acs2024_rename$`College Only`), 
          max(acs2024_rename$`Advanced Degree`),
          max(acs2024_rename$`Employed`), 
          max(acs2024_rename$`Unemployed`),
          max(acs2024_rename$`Not in Labor Force`), 
          max(acs2024_rename$`Commute Time (mins)`),
          max(acs2024_rename$`Total Income ($)`, na.rm = TRUE))) %>% 
  knitr::kable(caption = "Summary Statistics for 2024 ACS Sample",
               format = "latex", 
               booktabs = TRUE,
               digits = 2)
               
acs2024_stats



```

```{r}
acs2024_rename <- acs2024 %>%
  rename( "Age" = "AGE", 
     "Female" = "female" , 
     "Less than High School"= "less_than_hs" , 
     "High School Only"= "hs_only",
     "Some College" = "some_college" , 
    "College Only"= "college_only" , 
    "Advanced Degree" = "advanced_degree" , 
   "Employed" =  "employed" , 
     "Unemployed" = "unemployed" , 
     "Not in Labor Force" = "not_in_labor_force", 
     "Commute Time (mins)" = "TRANTIME", 
    "Total Income ($)" = "INCTOT")

library(modelsummary)
acs2024_table <- datasummary(
  `Age` + `Female` + `Less than High School` + `High School Only` + `Some College` + `College Only` +  `Advanced Degree` + `Employed` + `Unemployed` + `Not in Labor Force` + `Commute Time (mins)` + `Total Income ($)` ~ N + Mean + SD + Min + Max, 
  data= acs2024_rename, 
  title = "Summary Statistics for 2024 ACS Sample", 
   output = "latex"
  )


```


6.Does the maximum commute time make sense for a daily commute? Write 1–2 sentences
Yes, this makes sense because if the maximum is 195 minutes, this is approximately 3.25 hours spent on a commute, which can become tiresome, but if it is mandatory and one often gets caught in traffic, this can become a reality for many people. 


#Analysis Section 

```{r, include=FALSE}

acs2024$TRANTIME

sum(acs2024_raw$TRANTIME == 0, na.rm = TRUE)
length(acs2024_raw$TRANTIME)

acs2024_raw$TRANTIME
```

1.  2062945 people have 0 minute commute time, there are a total number of 3422888 people. 60.27% of total commuters do not have a commuter time at all.

2.  People who would not have a commuter time would be children, retirees, unemployed people, remote workers, and people who carpool to the work.

```{r, include=FALSE}
commuters <- acs2024_rename %>% 
  filter(`Commute Time (mins)` != 0)
#acs2024_commuters<- acs2024 %>% 
#    mutate(commuters = if_else(TRANTIME %in% c(0, 888), NA_real_, TRANTIME))





#commuters variable 
```

#are we allowed to use some of the suggestions he provided? 3. My criteria for the commuters variable was anyone with a commuter time greater than 0 and not inclusive of 888. This is because 000 signifies "N/A" and 888 is coded to "Suppress for data year 2023 for select PUMAs." Therefore, my eliminating these from being included in the commuters variable and reclassifying as N/A these values are unable to impact the data and the summary of the variable.

```{r}
#sum data where it is not 0 or 888
sum(acs2024$commuters !=0 & acs2024$commuters !=888, na.rm = TRUE)


```

#is that correct code to be using? 3. There are 1359943 commuters left, which matches the remaining numbers that do not equal 0 or 888, after you subtract 3,422,888-2,062,945 (subtracting the total numebr of individuals - those who do not have a commute time).

```{r}
summary(acs2024$commuters)
```

3b. The minimum is 1.00 minute. This does not seem very feasible, but if someone were to live next to where they work, this could be a plausible explanation for having a 1 minute commute time.

```{r}
library(knitr)

acs2024_rename <- acs2024 %>% 
  rename( "Age" = "AGE", 
     "Female" = "female" , 
     "Less than High School"= "less_than_hs" , 
     "High School Only"= "hs_only",
     "Some College" = "some_college" , 
     "College Only"= "college_only" , 
    "Advanced Degree" = "advanced_degree" , 
   "Employed" =  "employed" , 
     "Unemployed" = "unemployed" , 
     "Not in Labor Force" = "not_in_labor_force", 
     "Commute Time (mins)" = "TRANTIME", 
    "Total Income ($)" = "INCTOT")

acs2024_commuter <- data.frame(
  Variable = c("Age", "Female", "Less than High School", "High School Only",  "Some College", "College Only", "Advanced Degree", "Employed", "Unemployed", "Not in Labor Force", "Commute Time (mins)", "Total Income ($)"),
  N = c(length(commuters$`Age`),
        length(commuters$`Female`),
        length(commuters$`Less than High School`),
        length(commuters$`High School Only`),
        length(commuters$`Some College`),
        length(commuters$`College Only`),
        length(commuters$`Advanced Degree`),
        length(commuters$`Employed`),
        length(commuters$`Unemployed`),
        length(commuters$`Not in Labor Force`),
        length(commuters$`Commute Time (mins)`),
        length(commuters$`Total Income ($)`)),
  Mean = c(mean(commuters$`Age`),
           mean(commuters$`Female`), 
           mean(commuters$`Less than High School`),
           mean(commuters$`High School Only`), 
           mean(commuters$`Some College`),
           mean(commuters$`College Only`),
           mean(commuters$`Advanced Degree`), 
           mean(commuters$`Employed`),
           mean(commuters$`Unemployed`), 
           mean(commuters$`Not in Labor Force`),
           mean(commuters$`Commute Time (mins)`),
           mean(commuters$`Total Income ($)`, na.rm = TRUE)),
  SD = c(sd(commuters$`Age`),
         sd(commuters$`Female`), 
         sd(commuters$`Less than High School`),
         sd(commuters$`High School Only`), 
         sd(commuters$`Some College`),
         sd(commuters$`College Only`), 
         sd(commuters$`Advanced Degree`),
         sd(commuters$`Employed`), 
         sd(commuters$`Unemployed`),
         sd(commuters$`Not in Labor Force`), 
         sd(commuters$`Commute Time (mins)`),
         sd(commuters$`Total Income ($)`, na.rm=TRUE)),
  Min = c(min(commuters$`Age`),
          min(commuters$`Female`), 
          min(commuters$`Less than High School`),
          min(commuters$`High School Only`), 
          min(commuters$`Some College`),
          min(commuters$`College Only`), 
          min(commuters$`Advanced Degree`),
          min(commuters$`Employed`), 
          min(commuters$`Unemployed`),
          min(commuters$`Not in Labor Force`), 
          min(commuters$`Commute Time (mins)`),
          min(commuters$`Total Income ($)`, na.rm = TRUE)),
  Max = c(max(commuters$`Age`),
          max(commuters$`Female`), 
          max(commuters$`Less than High School`),
          max(commuters$`High School Only`), 
          max(commuters$`Some College`),
          max(commuters$`College Only`), 
          max(commuters$`Advanced Degree`),
          max(commuters$`Employed`), 
          max(commuters$`Unemployed`),
          max(commuters$`Not in Labor Force`), 
          max(commuters$`Commute Time (mins)`),
          max(commuters$`Total Income ($)`, na.rm = TRUE))) %>% 
  knitr::kable(caption = "Summary Statistics for 2024 ACS Sample",
               format = "latex", 
               booktabs = TRUE,
               digits = 2)
               
acs2024_commuter
               
```

```{r}
acs2024 <-acs2024_raw %>% 
  select("STATEFIP", "SEX", "AGE", "EDUC", "EMPSTAT","INCTOT","TRANTIME") %>% 
  mutate(female = if_else(SEX == 2, 1, 0), 
  less_than_hs = if_else(EDUC %in% c(1,2), 1, 0),
  hs_only = if_else(EDUC %in% c(3,4,5,6), 1, 0), 
  some_college = if_else(EDUC %in% c(7,8,9), 1, 0), 
  college_only = if_else(EDUC ==  10,1,0),
  advanced_degree = if_else(EDUC == 11, 1, 0)) |> 
  
  mutate(employed = if_else(EMPSTAT == 1, 1, 0, 
              if_else(EMPSTAT %in% c(0,9), NA_integer_, 0)),
  unemployed = if_else(EMPSTAT == 2, 1, 0, 
                if_else(EMPSTAT %in% c(0,9), NA_integer_, 0)), 
  not_in_labor_force = if_else(EMPSTAT == 3, 1, 0, 
   if_else(EMPSTAT %in% c(0,9), NA_integer_, 0))) |>
  mutate(INCTOT = if_else(INCTOT %in% c(9999999, 9999998), NA_real_, INCTOT)) %>% 
   mutate(commuters = if_else(TRANTIME %in% c(0, 888), NA_real_, TRANTIME))


acs2024_rename <- acs2024 %>% 
  rename( "Age" = "AGE", 
     "Female" = "female" , 
     "Less than High School"= "less_than_hs" , 
     "High School Only"= "hs_only",
     "Some College" = "some_college" , 
     "College Only"= "college_only" , 
    "Advanced Degree" = "advanced_degree" , 
   "Employed" =  "employed" , 
     "Unemployed" = "unemployed" , 
     "Not in Labor Force" = "not_in_labor_force", 
   "Commute Time (mins)" = "TRANTIME",
     "Commute Time (mins)-Analytical" = "commuters", 
    "Total Income ($)" = "INCTOT")

library(modelsummary)
acs2024_table <- datasummary(
  `Age` + `Female` + `Less than High School` + `High School Only` + `Some College` + `College Only` +  `Advanced Degree` + `Employed` + `Unemployed` + `Not in Labor Force` + `Commute Time (mins)` + `Total Income ($)` ~ N + Mean + SD + Min + Max, 
  data= acs2024_rename, 
  title = "Table 1: Summary Statistics for 2024 ACS Sample"
  )
```
